A Type System for R: Empirically Designed and Evaluated

### 2019-09-23 paper/main.tex
new vers

\documentclass[acmsmall,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{booktabs,listings,xspace,wrapfig}
\lstset{language=R}
\usepackage{my_style}
\definecolor{LightGray}{rgb}{.95,.95,.95}
\definecolor{Gray}{rgb}{.3,.3,.3}
\definecolor{DarkGray}{rgb}{.5,.5,.5}

@@

% Macros for type names?
\newcommand{\reals}{$\mathtt {\rm I\!R}$\xspace}
\newcommand{\realv}{$\overline{\rm I\!R}$\xspace}

\setcopyright{none}
%\setcopyright{acmcopyright}%\setcopyright{acmlicensed}
%\acmDOI{10.475/123_4}
%\acmConference[OOPSLAs]{Woodstock conference}{July 1997}{El Paso, Texas USA}
%\acmYear{1997}%\copyrightyear{2016}%\acmPrice{15.00}
\begin{document}

\title{A Large-Scale Study of Polymorphism in R}

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{20\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
\newcommand{\METAARGCOUNT}{7,051\xspace}
\newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}
\newcommand{\PERCSCALARMONO}{38.2\%\xspace}

\newcommand{\LZEROPERCPOLY}{19\%\xspace}
\newcommand{\LONEPERCPOLY}{18.1\%\xspace}
\newcommand{\LTWOPERCPOLY}{21.3\%\xspace}

\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
\newcommand{\attrclass}[3]{\ensuremath{#1^{\mathtt{#3}}_{\mathtt{#2}}}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
%\renewcommand{\l}{\ensuremath{\small L<?>}\xspace}     % List
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt{#1}}}\xspace}     % list<T>
\newcommand{\M}[1]{\ensuremath{\attr{\vec{\tt #1}}{mat}}\xspace}     % matrix
\newcommand{\df}{\ensuremath{\attr{\l}{df}}\xspace}     % data.frame

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  In order to answer
it, we study the polymorphism that is present in \RLOC lines of R
code spread among some \PACKAGES packages, written over a
period of over \YEARS years by thousands of programmers.  We perform a dynamic
analysis, leveraging tests and use-cases, to determine the level of
polymorphism that is present in the code. We do this for several potential
notions of types. Our results suggest that polymorphism is important in some
key parts of the system but that relatively simple type annotations could be
used to capture most of the interesting cases.

@@

\section{Introduction}

Our community builds, improves, and reasons about programming languages.  To
make design decisions that benefit most users, we need to understand the
language we are working with as well as the real-world needs it
answers. Often, we as researchers can appeal to our intuition, as many
languages are intended to be general purpose and attract users with some
computer science training. Unfortunately, these intuitions do not always
apply to domain-specific languages, languages designed for and by a specific
group of users to solve very specific needs. This is the case of the data
science language R.

R and its ancestor S are languages designed, implemented, and maintained by
statisticians. Originally they were designed as glue languages, languages
that would allow to read data into vectors and call statistical routines
written in Fortran. Over three decades, the languages became widely used
across many fields of science and in industry for data analysis and data
visualization; with time additional features were added.  Modern R, as a
linguistic object of study, is fascinating. It is a vectorized, dynamically
typed, lazy functional language with limited side-effects, extensive
reflective facilities and retrofitted object-oriented programming support.

Many of the design decisions that gave us R were intended to foster an
interactive, exploratory, programming style. These include, to name a few,
the lack of type annotations on variables and functions, the ability to use
syntactic shortcut, and the automatic conversion between data types.  While
these choices have led to a language that is surprisingly easy to use for
beginners---many data science programs do not teach the language itself but
simply introduce some of its key libraries---they have also created a
language where almost all computations yield a numeric result and where
errors can go undetected.

One way to increase assurance in the results obtained when using R would be
to add type annotations to functions and variable declarations. These
annotations could then be used, either statically or (more likely)
dynamically, to catch mismatches between expected and provided data values.
The nature of R is such that it is unlikely to be ever fully statically
checked, furthermore end users may not be willing to write types when
carrying out exploratory programming tasks. So, we are looking for an
optional type system that would allow us to capture as much of behavior of
library functions as possible while remaining easy to understand for
end-users and library developers alike.

This paper is a data-driven study of what a type system for the R language
could look like. Our intention is to propose changes to the
language, and for any changes to be accepted by the user community they
must clearly benefit the language without endangering backwards
compatibility. Our goal is thus to find a compromise between simplicity and
usefulness; the proposed type system should cover most common programming
idioms while remaining easy to use. In order to do this, we need to
understand the degree of polymorphism present in R code, that is to say, how
programmers leverage the dynamic nature of R to write code that can accept
arguments of different types.  This understanding will drive our design.

We propose to capture the degree of polymorphism present in R by means
of a dynamic analysis of widely used libraries. For each function call we
can record the types of its arguments and of its return value. This allows
us to observe how many different combination of types are accepted by any
given function. Unlike many other languages, R has a carefully curated
software repository called CRAN. To be deposited in CRAN, a package must
come with sample dataset, tests and executable use-cases. As part of normal
operations these tests are run regularly and failing packages are removed.
This allowed us to have access to \PACKAGES libraries and about an order of
magnitude more runnable scripts that exercise those libraries.

The contributions of this paper are thus as follows:
\begin{itemize}
\item A large-scale analysis of the polymorphism present in function
  signatures of \PACKAGES widely used and actively maintained R packages.
\item A tracing and analysis pipeline that extends a previously published
  test generation tool named \genthat.
\item A set of type annotations that captures how programmers use types
in existing R code.
\end{itemize}

One threat to validity of our work is that we rely on dynamic analysis, so
our conclusions are only as good as the coverage of the possibly function
behaviors. Previous work~\cite{issta18} reported that running all the
scripts that come with CRAN packages gives, on average, 68\% test coverage.
We attempted to mitigate the threat coming from the fact that only part of
the code is being exercised by manual analysis. It would be reasonable to
ask for confirmation of the data by static analysis of the code, but sound
static analysis of R is difficult because of the extensive use of reflective
features such as \code{eval} and of the ability to redefine the meaning of
operators such as \code{+} and \code{if}.  Another threat to validity is
that we only have access to code that has been deposited in the CRAN
repository. While this may bias our findings towards code written to be
reusable and, possibly, better engineered than typical user code. This is
also the code that would most benefit from type annotations.

\newpage  %%Leave here

\section{The R Programming Language}\label{sec:rlang}

Over the last decade, the R Project has become a key tool for implementing
sophisticated data analysis algorithms in fields ranging from Computational
Biology~\cite{R05} to Political Science~\cite{R:Keele:2008}. At the heart of
the R project is a \emph{vectorized, dynamic, lazy, functional,
  object-oriented} programming language with a rather unusual combination of
features~\cite{ecoop12} designed to ease learning by non-programmer and
enable rapid development of new statistical methods.  The language, commonly
referred to as R was designed in 1993 by Ross Ihaka and Robert
Gentleman~\cite{R96} as a successor to S~\cite{S88}.  First released in
1995, under a GNU license, R rapidly became the lingua franca for
statistical data analysis. Today, there are over 13,000 R packages available
from repositories such as CRAN and Bioconductor.  With 55 R user groups
world-wide, Smith~\cite{eco11} estimates that there are over 2 million
end-users.

As an introduction to R, consider the code snippet in Fig.~\ref{sample} from
a top-level interaction where the user defines a function \code{normSum}
that accepts vectors of integers, logicals, doubles and complex values and
normalizes the vector with respect to its sum and rounds the results. The
function definition does not require type annotations, and all operations
transparently work on vectors of any length and different types.

\begin{figure}[!hb]{\small
\begin{lstlisting}[style=R]
> normSum <- function( m )  round( m / sum(m), 2)
> normSum(c(1L,3L,6L))
[1] 0.1 0.3 0.6
> normSum(c(1.1,3.3,6.6))
[1] 0.1 0.3 0.6
> normSum(c(1.6,3.3,6.1))
[1] 0.15 0.30 0.55
> normSum(complex(r=rnorm(3),i=rnorm(3)))
[1] 0.49+0.21i 0.30-0.18i 0.22-0.03i
\end{lstlisting}}
\caption{Sample R code}\label{sample}
\end{figure}

In R, function can be called with named parameters, R support variable
argument lists, and arguments can have default values. Putting all of these
together consider the following declaration:

\begin{lstlisting}[style=R]
f <- function(x, ..., y=3) x + y
\end{lstlisting}

\noindent
Function \k{f} can be called with a single argument \code{f(3)}, with named
argument \code{f(y=4,x=2)} and with a variable number of arguments,
\code{f(1,2,3,4,y=5)}, all of these calls will return \code{6}.

R has a number of features that are not crucial to the present
discussion. We will mention some of them here for completeness.  In R, data
structures are reference counted and have copy-on-write semantics, thus the
assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
the reference count on that object is 1.  This semantics gives R a
functional flavor while allowing updating in place within loops (the first
update copies, subsequent updates are performed on the copy). Arguments to
functions are evaluated only when needed, they are bundled in so-called
promises which package the original expression (as an abstract syntax tree, or AST), its environment
as well as the result of evaluating the expression. Promises can be
leveraged for meta-programming as it is possible to retrieve the text of a
promise and evaluate it in a different environment.

\subsection{Types of Data}

Before attempting to define a type system for R, we should understand the
different kinds of values that programs operate on.  As we will see
different notions of type may emerge depending on how granular we want to
be.

\renewcommand{\k}[1]{{\tt #1}\xspace}

R has one builtin notion of type that can be queried by the \k{typeof}
function. Over the years, programmers have found the need for a richer type
structure and have added {\it attributes}. The best way to think of attributes is
as an optional map from name to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.

\begin{wrapfigure}{r}{6.1cm}
\footnotesize\begin{tabular}{l|c|l@{}}\hline
\multicolumn{3}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}   & \L & vector of boolean values\\
\k{integer}   & \I & vector of 32 bit integer values\\
\k{double}    & \D & vector of 64 bit floating points\\
\k{complex}   & \X & vector of complex values\\
\k{character} & \C & vector of strings values\\
\k{raw}       & \R & vector of bytes\\
\k{list}      & \l & vector of values of any type\\\hline
\multicolumn{3}{l}{\bf Scalar data types:}\\\hline
\k{NULL}      & \sN &  singleton null value\\
\k{S4}        & \sS &  instance of a S4 class \\
\k{closure}   & \sF & a function with its environment\\
\k{environment}&\sE &  a mapping from symbol to value \\\hline
\multicolumn{3}{l}{\bf Implementation data types:}\\\hline
\multicolumn{3}{l}{\k{special},
\k{builtin},
\k{symbol} (\sY),
\k{pairlist},
\k{promise}}\\
\multicolumn{3}{l}{
\k{language},
\k{char},
\k{...},
\k{any},
\k{expression},
}\\
\multicolumn{3}{l}{
\k{externalprt},
\k{bytecode},
\k{weakref}}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{wrapfigure}

Figure~\ref{types} lists all of the builtin types that are provided by the
language. They are the possible return values of \k{typeof}. There is no
intrinsic notion of subtyping in R. But, in many context a \k{logical} will
convert to \k{integer}, and an \k{integer} will convert to \k{double}.  Some
off conversion can occur in corner cases, such as \k{1<"2"} holds and
\k{c(1,2)[1.6]} returns the first element of the vector, as the double is
converted to an integer. R does not distinguish between scalars and vectors
(they are all vectors), so \code{typeof(5) ==} \code{typeof(c(5)) ==
  typeof(c(5,5))} \code{ == "double"}. Finally all vectorized data types have a
distinguished missing value denoted by \code{NA}. The default type of
\code{NA} is \k{logical}. We can see that \code{typeof(NA)=="logical"}, but
NA inhabits every type: \code{typeof(c(1,NA)[2])=="double"}.

With one exception all vectorized data types are monomorphic, the exception
is the \k{list} type which can hold values of any other type including
\k{list}. For all monomorphic data types, attempting to store a value of a
different type will cause a conversion. Either the value is converted to the
type of vector, or the vector is converted to the type of the value.

Scalar data types include the distinguished \k{NULL} value, which is also of
type \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types that are mostly not used by user code, they are listed in
Figure~\ref{types} for reference.

The addition of attributes lets programmers extend the set of types by
tagging data with user-defined attributes. For example, one could define a
vector of four values, \code{x<-c(1,2,3,4)} and then attach the attribute
\k{dim} with a pair of numbers as value: \code{attr(x,"dim")<-c(2,2)}.  From
that point, arithmetic functions will treat \k{x} as a 2x2 matrix. Another
attribute that can be set is the \k{class}.  This attribute can be bound to
a list of class names. For instance, \code{class(x)<-"human"}, set the class
of \k{x} to be \k{human}.  Attributes are thus used for object-oriented
programming. The S3 object system support single dispatch on the class of
the first argument of a function, whereas the S4 object system allows
multiple dispatch (on all arguments). Some of the most widely used data
type, such as data frames, leverage attributes. A data frame, for instance,
is a list of vectors with a class and a column name attribute.

\paragraph{Summary.} The most common values in R computations are vectorized
types. R programs do not have a way to constrain values to be scalar.
\k{NULL} is sometimes used to represent the case when no value is
available. \k{NA} is used within vector to represent missing observations.
Attributes can decorate values and are used as building blocks for
object-oriented programming. A potential type system for R could focus only
on the builtin types, if one wanted to strive for simplicity, or it could
try to capture attributes at the risk of increased complexity.

\newpage
\section{Corpus}\label{sec:corpus}

In this section, we present our dataset. The R language aims to accommodate
data analysts; their workflows start with data import, followed by cleaning,
and then by steps of modeling, transformation and visualization. Often, the
code of these analysis pipelines resides, together with the data and
results, in notebooks. Few notebooks are publicly shared, and when they are,
the data is not. For this reason our analysis focuses on packages which
bundle reusable units of R code with documentation, sample data and
use-cases.

We focus on packages hosted on the \emph{Comprehensive R Archive Network} or
CRAN.  With over 13,000 packages, CRAN is the largest repository of software
written in R. It is experiencing sustained growth with an average of size
new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
\emph{curated} collection: A package is only accepted to CRAN if it abides
by a number of well-formedness rules.  Most relevant for our purposes,
packages must have data, examples, vignettes and tests, all of which must
successfully run. From our perspective this means that each package in CRAN
comes with several executable scripts that exercise some of its
functionality.  Notable exceptions to this rule are packages only containing
data, which have no runnable code but are referenced by other packages.
Only \DATAPKGS packages had no executable code, accounting for \DATAPKGSPERC
of CRAN.

The corpus used in this paper is a subset of CRAN. We retained packages that
could be run by our infrastructure in less than one hour. This corpus
consists of \PACKAGES packages, accounting for some \PERCENTCRAN of all
packages.  These packages have a total of \RLOC lines of R code and \CLOC
lines of C code. Figure~\ref{allcloc} shows a per-package breakdown of the
size of each package sorted by increasing numbers of lines of R. The figure
suggests that there is little correlation between the C/C++ parts and how
many lines of R a package contains. The median size of a package is 541
lines of R code and the largest package has 86K LOC. Typically, C/C++ code
is used to implement performance critical portions of the code. The majority
of packages, 8,375 to be precise, have no C or C++ code.  The remaining 3,078
packages have a median 572 lines of C and the largest package has 385,839
lines of C.

\begin{figure}[!b]\begin{center}
\includegraphics[width=.9\textwidth]{linesofrandccode}
\caption{Lines of code (log scal); for each package, R is above 0, and C/C++
  below}\label{allcloc}\end{center}
\end{figure}

For each package, we extracted all executable code snippets from
documentation, vignettes and tests and ran them independently, recording all
calls to R functions.  It is noteworthy that in order to run the scripts in
one package, it is often necessary to load a number of other packages.
In~\cite{issta18}, the authors estimated code coverage to be around 68\%
when including reverse dependencies.  As our infrastructure adds overhead to
script execution, coupled with the fact that some scripts take inordinate
amounts of time to run, we limited our analysis of any one
package to one hour.

% https://www.r-pkg.org/downloaded
\begin{figure}[!th]{\footnotesize\begin{tabular}{@{}r||l|r|r|r|r|r@{}}\hline
\bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts & \bf Calls Observed & \bf Signatures Recorded \\
\hline
\tt Rcpp  & Seamless C++ integration & 2.2K & 4.2K & 25 & 55K & 340 \\
\tt rlang & Functions for 'Tidyverse' & 7.0K & 6.1K & 122 & 3,924K & 8,422 \\
\tt glue  & Interpreted string literals & 0.3K & 0.3K & 8 & 4K & 145 \\
\tt tibble & Simple data frames & 2.0K & 0.3K & 16 & 1,332K & 6,367 \\
\tt stringi &  String processing & 1.5K & 515K & 64 & 923K & 873 \\
\tt ggplot2 & Data visualisations & 14K & 0 & 130 & 153K & 4,608 \\
\tt dplyr  &  Data manipulation & 4.5K & 4.7K & 78 & 233K & 3,099 \\
\tt pillar & Formatting for columns & 1.4K & 0 & 13 & 803K & 1,514 \\
\tt R6 & Classes w. ref. semantics & 0.7K & 0 & 2 & 1K & 330 \\
\tt stringr & String operations & 0.5K & 0 & 32 & 1,764K & 534 \\
\end{tabular}}\caption{10 Most Downloaded Packages.}\label{most}
\end{figure}

Figure~\ref{most} shows the ten most downloaded CRAN packages.  For each
one, we list how many lines of R and C/C++ the packages contains.  We show
the number of scripts that could be extracted from the package. Each script
corresponds to either one use-case or a set of unit tests. We print the
number of function calls that were observed by our infrastructure, and the
number of unique call signatures that were recorded.

Our infrastructure only retains unique argument/return combinations. Thus,
while we observe large number of functions being called with different
values, the types of these function call are often similar. Over the entire
corpus, we can see the relation between observed and recorded calls in
Figure~\ref{recorded}.  The median number of observed calls is 82 and
maximum is 19 million.  The median number of recorded signatures is 16 and
the maximum is 8,422. These numbers are skewed by a number of scripts doing
very few calls before plunging into C code.

\begin{figure}[htbp]\begin{center}
\includegraphics[width=.9\textwidth]{recordsbypkg}
\caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
\label{recorded}\end{center}
\end{figure}

\newpage
\section{Methodology}

In this section, we detail our methodology for collecting data.  Our aim is
to observe arguments and return values of function calls, and from these
generalize possible type signatures for the called functions.  We base our
infrastructure on an open source tool called \genthat whose purpose is to
generate unit tests for R libraries~\cite{issta18}.  \genthat achieves this
by synthesizing unit tests from recorded function argument and return
values, comparing candidate tests against existing ones to avoid generating
tests which do not increase code coverage.  To suit our purposes, we change
the existing tool in two main ways: we record \emph{shapes} rather than
values, and ignore the code coverage optimization phase.  Both changes are
beneficial for scalability, allowing us to trace far more calls.

To illustrate our approach, consider the following script which adds a
double to an integer, and then creates a matrix from a vector, finally
adding them together to get a new matrix.

\begin{figure}[!hb]
\begin{tabular}{ll}\begin{minipage}{5cm}
{\small\begin{lstlisting}[style=R]
> 1 + 2L
[1] 3
> x <- c(1,2,3,4)
> attr(x,"dim") <- c(2,2)
> x + c(1,2)
     [,1] [,2]
[1,]    2    4
[2,]    4    6
\end{lstlisting}}
\end{minipage} &
\begin{minipage}{8cm}\small
\begin{tabular}{rl}
\tt `+`: &\tt \sD \sI \to \sD \\
\tt c:& \sD \sD \sD \sD \to \D\\
\tt `<-`: &\tt \sY \D \to \D\\
\tt c: & \tt \sD \sD \to \D\\
\tt `attr<-`:&\tt  \sY \C \D \to \\
\tt c: &\tt \sD \sD \to \D\\
\tt `+`: &\tt \attr\D{dim} \D \to \attr\D{dim}
\end{tabular}
\end{minipage}
\end{tabular}
\caption{Example script and recorded signature}\label{example}\end{figure}

The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
\k{`attr<-`}, and the shapes we expect to record are given to the right of the figure.
We abbreviate the types \k{double}, \k{integer}, \k{symbol} and \k{character} to \sD, \sI, \sY, and \sC for simplicity.
\D, \I, \Y, and \C denote vectors, whereas \sD, \sI, \sY, and \sC denote scalars.
We record the lengths of all vectorized-typed values, and for all types we record their classes and attribute names. 
Looking at the signatures observed for addition, it is clear that the function is polymorphic: it starts with the addition of two scalar numbers of different types, and later adds a matrix to a vector returning a matrix.

\subsection{Implementation}
\label{sec:Impl}

A high-level description of the workflow of our tool for one package
retrieved from CRAN is as follows:

\begin{enumerate}
\item {\bf Exec generation:} All runnable code in the package is extracted
  from its tests, examples and vignettes. The code snippets are combined
  into a single file.
\item {\bf Installation:} All packages that are required for execution of
  the current package are downloaded and installed.
\item {\bf Instrumentation:} As code is loaded into the R, every function
  definition is instrumented with an \code{on-exit} hook which is invoked
  when the function returns either normally or through an exception.
\item {\bf Recording:} When a hook is called, arguments and return value are
  inspected. We record \k{typeof}, \k{class} and \k{attributes} recursively.
  For \code{list} values, an extra bit of analysis is performed to record
  the element types.
\item {\bf Writing:} Unique signatures are recorded to file with information
  about which package triggered the recording.
\end{enumerate}

In R, all arguments are passed as promises, and unused arguments will remain unevaluated promises, which we record as {\it unevaled}. 
Our recording mechanism captures the order in which arguments are passed, can figure out which arguments were {\it not} provided (recording them as {\it missing} if the missing argument has no default value), and does {\it not} force unforced promises.
In R, trying to use missing values results in an error and thus an exception, triggering our recording phase: arguments which cause errors in this way are recorded as {\it error}.
For practical reasons, we do not record the contents of environments:
These can be used as hash tables, may be large, and are quite likely different from one another.

Unevaluated arguments in R are sometimes {\it carefully} used by their functions to implement some form of meta-programming.
Consider the {\tt magrittr} package, which implements a pipe function \%$>$\% that implements a flavor of function currying:

{\small\begin{lstlisting}[style=R]
mul <- function(x,y)  x*y
4 %>% mul(3)
\end{lstlisting}}

The call to the pipe desugars to \code{`\%>\%`(lhs=4, rhs=mul(3))} where the
\code{rhs} is not a valid expression. It is treated as code that will
manipulated to add one argument (the \code{lhs}), resulting in the function
actually performing a call of the form \code{mul(4, 3)}.

As we alluded to at the beginning of this section, our analysis collects shape information about
arguments and return values in order to be as specific as possible. The details of the information we collect follows.
Note that the first step of any type collection in our analysis begins with \code{typeof}.

\begin{itemize}
\item If the analysis encounters a primitive vector, say
  \code{typeof(x)=="double"}, it records the length of {\tt x} to determine
  if it can be treated as a scalar.
\item For a list, \code{typeof(x)=="list"}, shape information on all
  elements is used to describe the list's shape.  To avoid undue slowdowns,
  we only collect the content's \code{typeof}, and ascribe \ANY (any) if the
  elements are not of the same kind.
\item If the analysis encounters a scalar \code{NA}, we ascribe a unique
  NULL type.  In R, \code{NA} inhabits all types, but for a scalar NA,
  \code{typeof(NA)=="logical"}. Scalar NAs are used as uninitialized values.
\item For a matrix, \code{typeof(m)=="some\_primitive\_type"}, and\code{m} will have matrix class and also a dims and
  optionally dimnames attribute. Here, we ascribe the shape \attr{\tt T}{mat}
  where {\tt T} is the primitive element type.
\item For a data frame, \code{typeof(df)=="list"}, and \code{df} will have class {\tt data.frame}, and {\tt names} and {\tt row.names} attributes.
 To reflect this, we ascribe the shape \attr{\l}{df}.
\end{itemize}

The infrastructure memoizes emitted signatures to avoid writing multiple
identical signatures for the same function.

\section{Types for R}

Our aim is to capture the polymorphism in R, and to do so we must define a
meaningful set of basic types for the language. Also, we are interested in
finding a type system that could be adopted, for this we aim to investigate
the simplest possible solution.  There are several possible choices, with a
variety of levels of expressivity and complexity.  For concreteness, we will
pick out three different sets of types.  To start simple, we define a set of
types which closely align with R's builtin types, and later define richer
notions of type which incorporate attributes, classes, and capture the
various object systems of R.

Throughout this paper, we will say that an argument is {\it polymorphic at}
\textbf{\emph{LX}} (read: level X) if it has been called with at least two
values that belong to distinct types at {\bf LX}.  \textbf{L2} will
incorporate class and attribute lists, which are slightly trickier to
``type'', and we defer discussion of this until Section~\ref{sec:L2def}.

\subsection{L0 Types}
\label{sec:L0def}

The logical ``baseline'' for the design of a type system for R is to adopt
the language's builtin types.  We dub these {\bf L0} types, and they are
primarily built on the \code{typeof} values, which reflects the
implementation of the type in the runtime environment.  More specifically,
the \code{typeof} a value is simply the runtime type tag of the value.
Given that, {\bf L0} includes the following types: vectors of primitive
types (integers \I, double \D, character \C, logical \L, raw \R, complex
\X), lists of any type (\l), and scalar values (environment \sE, closures
\sF, symbols \sY, S4 object \sS, and null \sN). We omit short forms for some
of R's more esoteric types (e.g., {\tt environment}s).  We add the any
(written \ANY) type to stand in for unevaluated values Subtyping in {\bf L0}
follows from the conversion rules of the language: Logicals are subtypes of
integers, which are themselves subtypes of doubles (so \L <: \I <: \D).  All
types are subtypes of \ANY, thus {\tt T}<:\ANY for an type {\tt T}.

%
\subsection{L1 Types}
\label{sec:L1def}

{\bf L0} is limited in expressivity as it (1) supports polymorphism for
lists in a trivial way, (2) does not address some of the popular constructed
data types such as matrices and data frames, and (3) cannot differentiate
between scalar values and vectors.  {\bf L1} types build on and extend {\bf
  L0} with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
and \sX), type-parametric lists (written \lT{T} for any type {\tt T}), data
frames (written \attr{\l}{df} as they are built up from lists of type any),
and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
type {\tt T}. The subtype relation is undefined for lists of different
types, as well as for matrices and data frames.

\subsection{L2 Types}\label{sec:L2def}

{\bf L2} extends the previous system with support for classes and
attributes.  While there are a finite number of results from the
\code{typeof} function, this is {\it not} the case for \code{class} and
\code{attributes}, as classes and attributes {\it can} be user-defined.  In
terms of classes, R has a number of built-in or otherwise standard classes,
ranging from truly primitive classes such as {\tt numeric} to the more
complex {\tt data.frame}.  {\bf L1} already incorporates these builtin
classes, namely through its data frame and matrix types, so we will ignore
those classes and thus highlight those which are user-defined.  As for
attributes, again some patterns arise naturally in R, e.g. all matrices have
a {\tt dims} attribute, but even these are somewhat interesting, as a
programmer could in theory set their own {\it dims} attribute on an
arbitrary value.  The only attribute we ignore is the {\tt class} attribute,
as we collect the class of values separately (and the {\tt class} attribute,
if present, is always equal to the class of a value).  Attributes are
name-value pairs, and to keep things simple we consider only attribute
names.

In sum, we extend {\bf L1} with attribute lists and classes:
\attrclass{T}{C}{A} where {\tt T} can be any L1 type other than data frame
or matrix, and {\tt C} and {\tt A} are lists of classes and attribute names
respectively.  The subtype relation is defined as: \attrclass{T}{C}{A}
$<:_{\mathbf{L2}}$ \xspace \attrclass{S}{C'}{A'} if $ T
<:_{\mathbf{L1}} S$ (i.e. are subtypes in the sense of {\bf L1}), $C
\subseteq C'$, and $A \subseteq A'$.

\subsection{Simplification by Subtyping}\label{sec:autosub}

When analyzing our data, we perform a certain amount of simplifications
automatically to minimize the footprint of equivalent signatures.  This
subtyping is performed at the function level of granularity.  For example,
if a function is observed to have two signatures \sD \to \df to \D \to \df,
we collapse the signature to \D \to \df as \sD $<:$ \D and \df $<:$ \df.  We
would not collapse \sD, \I \to \df and \D, \sI \to \df, as \I $\not <:$ \sI
and \D $\not<:$ \sD.

%
%
%
\section{Examples of Polymorphism}\label{sec:polyex}

Before presenting the results of our analysis, we  quickly survey
examples of real, polymorphic R functions that we have analyzed, to build up
a global sense of how types manifest themselves in current R code.

We begin by referring the reader to Figure~\ref{fig:realex}, which showcases
the compatibility of integers and doubles.  Coercion from integer to double
is performed automatically where appropriate, and likewise coercion from
double to integer is automatically performed when integers are strictly
required.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> 5L + 1L # => 6L, an integer
> 5L + 1.2 # => 6.2, a double (5L coerced to 5.0)
> c(10, 20, 30)[1.2] # => 10, 1.2 coerced to 1L
\end{lstlisting}}\caption{Example of \sD, \sI type usage}\label{fig:realex}\end{figure}

Figure~\ref{fig:optnull} shows a function with an optional NULL argument.
In the function, {\tt x} is a sorted vector, and {\tt w} a vector with
default value NULL.  If {\tt w} is NULL, then default unit weights are
generated.  This pattern of default NULL (or NA) arguments is prevalent.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
frac.ranks <- function(x, w=NULL) {
  if (is.null(w)) w <- rep(1, length(x)) # if no weights passed, take all weights = 1
  ...
\end{lstlisting}}\caption{Optional argument ({\tt acid} package)}\label{fig:optnull}\end{figure}

Figure~\ref{fig:listvec} shows a function with list and vector polymorphism.
It takes a list or a vector {\tt point} and a data frame {\tt polyg}.  The
{\tt point} is explicitly converted to a vector (by \code{as.numeric})
regardless of type.  This showcases the ease of going from lists to vectors
(and conversely).

>>>>

\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage{booktabs,listings,xspace,wrapfig}
\lstset{language=R}
\usepackage{my_style}
\definecolor{LightGray}{rgb}{.95,.95,.95}
\definecolor{Gray}{rgb}{.3,.3,.3}
\definecolor{DarkGray}{rgb}{.5,.5,.5}

@@

\setcopyright{none}
\begin{document}

\title{A Type System for R: Empirically Designed and Evaluated}

\begin{abstract}

\AT{PLDI TODO:}

\item complete propagatr to generate constraints

\item feed existing data through resulting constraints

\item finish the annotation and contract generation code

\item write the paper

@@

%
%
%
%
%
%
\section{Introduction}

\begin{itemize}
  \item Adding types to dynamic languages is increasingly common.
  \item Types help write better code.
  \item We want to design a type system for R that is based on the patterns that we observe in current use of the language. 
  \item We also present a tool which takes a given package and synthesizes type signatures for package functions.
  \item This tool is tested by synthesizing annotations from package tests, examples, and vignettes, and running the code of the package's reverse dependencies (i.e. clients) to validate the type signatures.
  \item We find X, Y, Z.
\end{itemize}

The contributions of this paper are:

\item our analysis data;
\item a type system based on this data;
\item a tool to infer types from package code, and;
\item an evaluation of the tool on \AT{all of} CRAN.
